---
title: "The Effect of Transmission on MPG"
author: "Steve Myles"
date: "21 November 2015"
output:
  pdf_document:
    toc: no
---

# Executive Summary

This report is an investigation of the relationship between transmission (manual 
vs. automatic) and miles per gallon (MPG) for thirty-two automobiles featured in 
*Motor Trend* in 1974.  Specifically, the following two questions are addressed:

1. Is an automatic or manual transmission better for MPG?
2. Quantify the MPG difference between automatic and manual transmissions.

From the data, it is apparent that manual transmissions are better for MPG 
(on average, manual transmissions have higher MPG than automatic transmissions). 
The difference between them, removing confounding variables weight and 
horsepower, is approx. 2 MPG (average automatic transmission MPG = 34.00 and 
average manual transmission MPG = 36.09).

# Analysis

First, a brief exploratory analysis was performed.  This includes the generation 
of a box and whisker plot to compare the distributions of MPG for each type of 
transmission (see Appendix, Fig. 1).
```{r load and explore, echo = FALSE}
## load the data and provide a summary
data(mtcars)

## "am" is the field that contains transmission data.  0 = automatic 
## transmission and 1 = manual transmission.  convert binary variables am and 
## vs, as well as factor variables gear, carb, and cyl to factors
mtcars2 <- mtcars
mtcars2$am <- as.factor(mtcars$am)
mtcars2$vs <- as.factor(mtcars$vs)
mtcars2$gear <- as.factor(mtcars$gear)
mtcars2$carb <- as.factor(mtcars$carb)
mtcars2$cyl <- as.factor(mtcars$cyl)
```

From this plot, it is clear that on average and in the majority of cases (due to
the small overlap of the ranges) manual transmissions result in higher (better) 
MPG.  Additionally, the spread of MPG for manual transmissions is larger than 
that for automatic transmissions.

The analysis was continued with a simple linear model, comparing MPG to 
transmission: 

```{r regression 1, echo = FALSE}
## fit a linear model comparing mpg to transmission
modelmpgam <- lm(mpg ~ am, data = mtcars)
summary(modelmpgam)$coef
modelmpgamr2 <- summary(modelmpgam)$adj.r.squared
```

"Estimate" in the above table is the mean MPG for each factor, either automatic
(17.147, the intercept) or manual (24.39 = 17.147 + 7.245).  This confirms what 
is shown in the plot:  manual transmissions do, on average, result in higher MPG
(and are therefore better in this regard).  The note about the spread of values 
is also confirmed as MPG for manual transmissions has a higher standard error.  
This model's residuals plots can be seen in the Fig. 2 of the Appendix.

However, it should be noted that some of the other variables in the data set may
have a confounding effect.  This is evidenced by the model's relatively low 
adjusted *R^2^* (`r modelmpgamr2`, indicating that only 33.8% of the variation 
in MPG is explained by the variables in the model).  As such, it was necessary 
to fit other linear models that remove the effects of other variables from 
consideration in order to determine the true MPG difference between automatic 
and manual transmissions.

Per the following table which shows the correlations of `mpg` and the remaining 
nine variables (all variables in the `mtcars` data set with the exception of 
`am` and `mpg`), it is obvious that several other variables are related to MPG.

```{r mpg correlation, echo = FALSE}
## determine the correlation of mpg with each of the variables in the mtcars
## dataset, excepting am
mpgCor <- as.matrix(cor(mtcars[,1],mtcars[,c(2:8,10:11)]))
mpgCor
```

For the purposes of this analysis, only those variables with very strong 
(positive or negative) correlations were considered.  Very strong correlations
are defined here as +/- 0.7 (definition from [here](http://study.com/academy/lesson/positive-correlation-in-psychology-examples-definition-quiz.html)).

This leaves four variables, `cyl`, `disp`, `hp`, and `wt`.  The correlation of 
these variables with each other was then checked to avoid having the final model
include variables that are very strongly correlated with each other.

```{r other correlations, echo = FALSE}
## calculate the correlation matrix for all mtcars variables except mpg and am
mtcarsCor <- as.matrix(cor(mtcars[,c(2:4,6)]))
mtcarsCor
```

As can be seen in the above table, each of these variables has a very strong 
correlation with the others (using the +/- 0.7 definition as described above), 
with the  exception of horsepower (`hp`) and weight in thousands of pounds 
(`wt`).  As such, only these two variables were considered for inclusion into 
the linear model.

```{r regression2, echo = FALSE}
## fit additional linear models adding variables to isolate the transmission
## effect on mpg
modelmpgamwt <- lm(mpg ~ as.factor(am) + wt, data = mtcars)
modelmpgamwtr2 <- summary(modelmpgamwt)$adj.r.squared
modelmpgamwthp <- lm(mpg ~ as.factor(am) + wt + hp, data = mtcars)
modelmpgamwthpr2 <- summary(modelmpgamwthp)$adj.r.squared
modelmpgamwthpcoef <- summary(modelmpgamwthp)$coef
```

First, `wt` was added to the linear model resulting in an adjusted *R^2^* of 
`r modelmpgamwtr2`. This is much improved over the original adjusted *R^2^* of 
`r modelmpgamr2`.  `hp` was then added to the model, resulting in an adjusted
*R^2^* of `r modelmpgamwthpr2` (indicating that 82.3% of the variation in MPG
is explained by the variables in the model).  As this has the highest adjusted 
*R^2^* of the models considered, this is the best representation of the effect 
of transmission type on MPG.

The model's coefficients are:

```{r regression 2 coefficients, echo = FALSE}
modelmpgamwthpcoef
```

"Estimate" in the above table is the effect of each variable on mean MPG, either
automatic transmission (34.00, the intercept) or manual transmission (36.09 = 
34.003 + 2.084).  This confirms the earlier assessment that manual transmissions
on average result in higher MPG, though the difference between them is much less 
pronounced when the confounding effects of weight and horsepower are removed. 
Weight and horsepower both have negative effects on mean MPG, which makes 
logical sense - more powerful (higher horsepower) and/or heavier (higher weight) 
engines would require more fuel to operate and thus would have lower gas mileage 
(MPG). This model's residuals plots can be seen in the Fig. 3 of the Appendix.

# Note

This report was written in [R Markdown][rmd] and converted to pdf via [knitr][knitr]. 
For the source .Rmd document, please [click here][src].

[rmd]: http://rmarkdown.rstudio.com/
[knitr]: http://knitr.com/
[src]: https://github.com/scumdogsteev/regression_models_PA/blob/master/regression_models_PA.Rmd

\newpage

# Appendix

**Fig. 1** 

```{r boxplot, echo = FALSE}
## create a box and whisker plot of the relationship between these two variables
boxplot(mpg ~ am, data = mtcars, ylab = "MPG", xaxt = "n", 
        xlab = "Transmission", main = "Transmission and MPG Distributions for 
        Motor Trend 1974 Data")
axis(1, at=1:2, labels = c("Automatic", "Manual"))
```
\newpage

**Fig. 2** 

```{r modelmpgam plots, echo = FALSE}
## display residuals plots for modelmpgam
par(mfrow = c(2, 2))
plot(modelmpgam, main = "Transmission vs. MPG")
```
\newpage

**Fig. 3** 

```{r modelmpgamwthp plot, echo = FALSE, warning = FALSE, message = FALSE}
## display residuals plots for modelmpgamwthp
par(mfrow = c(2, 2))
plot(modelmpgamwthp, main = "Transmission vs. MPG 
with Confounding Variables Removed")
```